The Transition from Massive Metal Free Stars to Dwarf Galaxies

### 2011-01-19 ms.tex
Adding star formation history plot.
Finished first draft of the intro.

\title[High Redshift Dwarf Galaxies]{Metal Enrichment and Stellar
  Populations of High Redshift Dwarf Galaxies}

\author[Wise et al.]{John H. Wise$^1$\thanks{Hubble Fellow; e-mail:
    jwise@astro.princeton.edu}, Matthew J. Turk$^2$, Michael
  L. Norman$^2$, Tom Abel$^3$\\
  $^{1}$ Department of Astrophysical Sciences, Princeton University,
  Princeton, NJ 08544, USA\\
  $^{2}$ Center for Astrophysics and Space Sciences,
    University of California at San Diego, La Jolla, CA 92093, USA\\
  $^{3}$ Kavli Institute for Particle Astrophysics and Cosmology,
  Stanford University, Menlo Park, CA 94025, USA}

@@

The first generation of stars, so-called Population III, in the
universe were metal-free, forming from a primordial mixture of mainly
hydrogen and helium.  Heavy elements and dust grains were not present
in these objects to provide an efficient cooling mechanism in the
collapse of the host molecular cloud, whose collapses primarily rely
on molecular hydrogen line cooling in the gas-phase.  Without any
other cooling channels, the temperatures cannot cool below $\sim$ 200
K, resulting in Jeans masses on the order of 100 \Ms.  Numerical
calculations in the past decade have shown that Population III stars
have characteristic masses between 30 and 300
\Ms~\citep[e.g.][]{ABN02, Bromm02_P3, OShea07a}, which form in dark
matter (DM) haloes with masses $\sim 10^6 \Ms$
\citep[e.g.][]{MacLow86, Shapiro87, Tegmark97}.

These stars have large luminosities on the order of 10$^6 \lsun$,
surface temperatures of $\sim 10^5$ K, and lifetimes of $\sim3$ Myr
\citep{Bond84, Schaerer02}.  Radiation during their main sequence
creates an \hii region with $r \sim 1-3~\mathrm{kpc}$, ionizing the
surrounding intergalactic medium as the virial radii of the host haloes
are $r_{\rm vir} \sim 100 \mathrm{pc}$ \citep{Whalen04, Kitayama04,
  Alvarez06, Abel07}.  The over-pressurized \hii region drives a
$\sim$ 30 \kms~shock, which is 10 times greater than the escape
velocity of a $10^6 \Ms$ halo.  In turn, most of the gas is expelled
from the halo, leaving behind a warm ($3 \times 10^4$ K) and diffuse
(0.5 \cubecm) medium.  In haloes that are encompassed by these \hii
regions and not completely photo-evaporated \citep{refs}, the
additional free electrons catalyze both \hh~and HD molecular line
cooling \citep{OShea05, Johnson06, Yoshida07, McGreer08}, cooling the
gas below $\sim$ 100 K and lowering the Jeans mass to $\sim$ 10--30
\Ms.  These lower-mass, metal-free stars are called Population III.2,
whereas their more massive counterparts that form in unheated regions
are called Population III.1 stars \citep{Norman08}.

A fraction of Population III stars enrich the surrounding IGM when
they go supernova, which can happen in stars $\lsim 40~\Ms$ in Type II
supernovae (SNe) or in stars roughly between 140 \Ms~and 260 \Ms~in
pair-instability SNe \citep[PISNe;][]{2002ApJ...567..532H}. The host
halo and the neighboring haloes are then enriched with this ejecta.
The blastwave expands to a radius of a few kpc \citep{Bromm03_SN,
  Wise08_Gal, Greif10} with approximately half of the ejecta falling
back into the adjacent filaments and halos.  The blastwave may induce
star formation in nearby haloes through the compression of the gas
\citep{Ferrara98, more}, and the timescales for metal mixing into the
dense gas are many dynamical times \citep{Cen08} for shock velocities
$\lsim 100~\kms$.

When star-forming gas reaches some `critical metallicity' $Z_{\rm
  cr}$, radiative cooling from metal fine-structure lines or
\hh~formation on dust grains becomes efficient, allowing cooling to $T
\lsim 100$ K \citep{Omukai??}.  The exact value of $Z_{\rm cr}$ has
been constrained to be between $10^{-6} \zsun$ and $10^{-3.5} \zsun$.
The lower limit applies if dust cooling is important in these
scenarios \citep{Omukai05, Schneider06_Frag, clark08}, and the upper
limit happens if metal line cooling in the gas-phase is the dominant
process \citep{Bromm01, 2009ApJ...691..441S}.  This metal-enriched gas
fragments and most likely forms stars with masses similar to
present-day stellar initial mass functions (IMF), marking the local
transition from Population III to II star formation.  Intermediate to
these populations and possibly unique to high redshift, there exists a
mode of massive star formation that has a characteristic mass of $\sim
10 \Ms$ \citep{Larson98, Tumlinson07_IMF, 2009ApJ...691..441S}.  This
happens when cooling in metal-enriched gas is limited to the cosmic
microwave background (CMB) temperature $T_{\rm CMB} = 2.73 (1+z)$ K,
which can be significantly higher than the cores of typical molecular
clouds ($T \sim 10$ K).

The transition from Pop III to Pop II star formation is solely
dependent on the propagation of metals from the SNe remnants into
future sites of star formation.  Their flows are complex because of
the interactions between the SN blastwave, cosmological accretion and
halo mergers, and nearby stellar feedback.  \textbf{Discuss prior work
  on transition with the papers given by the previous referee.}

@@

Numerical simulations are useful to detangle and study these
complexities and the transition from Pop III to II stars.  Here, we
present simulations that include both modes of star formation and
their radiative, mechanical, and chemical feedback.  The methods used
here incorporate and link together recent results from metal-enriched
and metal-free star formation, the critical metallicity, and
pair-instability supernovae.  There are some uncertainities in these
input parameters, and we thus run several simulations while varying
these parameters.  This is the first time it has been possible to link
the formation and feedback of the first stars to protogalaxies in a
numerical simulation, resolving the important scales and including the
most important physical effects.

This paper is structured as follows.  Section \ref{sec:setup}
describes our simulation setup, numerical methods for star formation
and feedback, and different physical models considered in each
simulation.  Next we present our results on the stellar populations
and metal enrichement of the high-redshift dwarf galaxies in Section
\ref{sec:results}.  Then in Section \ref{sec:models}, we show the
changes in galactic properties and metal enrichments in each physical
model.  In Section \ref{sec:compare}, we compare our results with
previous work on the transition to Population II.  We then discuss the
importance of Population III star formation on high-redshift galaxy
formation and the implications on the interpretation of damped
Lyman-$\alpha$ absorbers (DLAs) and galactic archaeology in Section
\ref{sec:discuss}.  Finally in Section \ref{sec:summary}, we summarize
our results.

@@

%\begin{figure}
%  \plotone{sfh.eps}
%  \caption{\label{fig:SFR} Star formation history.}
%\end{figure}

>>>>

\title[Transition from Pop III to Dwarf Galaxies]{The Transition from
  Massive Metal Free Stars to Dwarf Galaxies}

\author[J. H. Wise et al.]{John H. Wise$^1$\thanks{Hubble Fellow;
    e-mail: jwise@astro.princeton.edu}, Matthew J. Turk$^2$, Michael
  L. Norman$^2$, Tom Abel$^3$\\
  $^{1}$ Department of Astrophysical Sciences, Princeton University,
  Princeton, NJ 08544, USA\\
  $^{2}$ Center for Astrophysics and Space Sciences,
  University of California at San Diego, La Jolla, CA 92093, USA\\
  $^{3}$ Kavli Institute for Particle Astrophysics and Cosmology,
  Stanford University, Menlo Park, CA 94025, USA}

@@

The first generation of stars, so-called Population (Pop) III, in the
universe were metal-free, forming from a primordial mixture of mainly
hydrogen and helium.  Heavy elements and dust grains were not present
in these objects to provide an efficient cooling mechanism in the
collapse of the host molecular cloud, whose collapses primarily rely
on molecular hydrogen line cooling in the gas-phase.  Without any
other cooling channels, the temperatures cannot cool below $\sim$ 200
K, resulting in Jeans masses on the order of 100 \Ms.  Numerical
calculations in the past decade have shown that Pop III stars
have characteristic masses between 30 and 300
\Ms~\citep[e.g.][]{ABN02, Bromm02_P3, OShea07a}, which form in dark
matter (DM) haloes with masses $\sim 10^6 \Ms$
\citep[e.g.][]{MacLow86, Shapiro87, Tegmark97}.

These stars have large luminosities on the order of 10$^6 \lsun$,
surface temperatures of $\sim 10^5$ K, and lifetimes of $\sim3$ Myr
\citep{Bond84, Schaerer02}.  Radiation during their main sequence
creates an \hii region with $r \sim 1-3~\mathrm{kpc}$, ionizing the
surrounding intergalactic medium as the virial radii of the host
haloes are $r_{\rm vir} \sim 100 \mathrm{pc}$ \citep{Whalen04,
  Kitayama04, Alvarez06, Abel07}.  The over-pressurized \hii region
drives a $\sim$ 30 \kms~shock, which is 10 times greater than the
escape velocity of a $10^6 \Ms$ halo.  In turn, most of the gas is
expelled from the halo, leaving behind a warm ($3 \times 10^4$ K) and
diffuse (0.5 \cubecm) medium.  In haloes that are encompassed by these
\hii regions and not completely photo-evaporated \citep{Shapiro04,
  whalen08}, the additional free electrons catalyze both \hh~and HD
molecular line cooling \citep{OShea05, Johnson06, Yoshida07,
  McGreer08}, cooling the gas below $\sim$ 100 K and lowering the
Jeans mass to $\sim$ 10--30 \Ms.  These lower-mass, metal-free stars
are called Pop III.2, whereas their more massive counterparts
that form in unheated regions are called Pop III.1 stars
\citep{Norman08}.

A fraction of Pop III stars enrich the surrounding IGM when
they go supernova, which can happen in stars $\lsim 40~\Ms$ in Type II
supernovae (SNe) or in stars roughly between 140 \Ms~and 260 \Ms~in
pair-instability SNe \citep[PISNe;][]{2002ApJ...567..532H}. The host
halo and the neighboring haloes are then enriched with this ejecta.
The blastwave expands to a radius of a few kpc \citep{Bromm03_SN,
  Wise08_Gal, Greif10} with approximately half of the ejecta falling
back into the adjacent filaments and halos.  The blastwave may induce
star formation in nearby haloes through the compression of the gas
\citep{Shapiro87, Ferrara98, Mackey03}, and the timescales for metal
mixing into the dense gas are many dynamical times \citep{Cen08} for
shock velocities $\lsim 100~\kms$.

When star-forming gas reaches some `critical metallicity' $Z_{\rm
  cr}$, radiative cooling from metal fine-structure lines or
\hh~formation on dust grains becomes efficient, allowing cooling to $T
\lsim 100$ K \citep{Omukai05}.  The exact value of $Z_{\rm cr}$ has
been constrained to be between $10^{-6} \zsun$ and $10^{-3.5} \zsun$.
The lower limit applies if dust cooling is important in these
scenarios \citep{Omukai05, Schneider06_Frag, clark08}, and the upper
limit happens if metal line cooling in the gas-phase is the dominant
process \citep{Bromm01, 2009ApJ...691..441S}.  This metal-enriched gas
fragments and most likely forms stars with masses similar to
present-day stellar initial mass functions (IMF), marking the local
transition from Pop III to II star formation.  Intermediate to
these populations and possibly unique to high redshift, there exists a
mode of massive star formation that has a characteristic mass of $\sim
10 \Ms$ \citep{Larson98, Tumlinson07_IMF, 2009ApJ...691..441S}.  This
happens when cooling in metal-enriched gas is limited to the cosmic
microwave background (CMB) temperature $T_{\rm CMB} = 2.73 (1+z)$ K,
which can be significantly higher than the cores of typical molecular
clouds ($T \sim 10$ K).

The transition from Pop III to Pop II star formation is solely
dependent on the propagation of metals from the SNe remnants into
future sites of star formation.  Their flows are complex because of
the interactions between the SN blastwave, cosmological accretion and
halo mergers, and nearby stellar feedback.  Spurred by the notion of a
critical metallicity, this transition has been extensively studied
with (i) volume-averaged semi-analytic models \citep{Scannapieco03,
  Yoshida04, Furlanetto05_Reion}, (ii) models using hierarchical
merger trees \citep{Tumlinson06, Tumlinson10, Salvadori07, Komiya10},
(iii) post-processing of cosmological simulations with blastwave
models \citep{Karlsson08, Trenti09, Trenti10}, and (iv) direct
numerical simulations with stellar feedback \citep{Tornatore07,
  Ricotti08, Maio10_Pop32, Maio10_Enrich}.

Over the past decade, these works have refined the general picture of
inhomogenous metal enrichment and the transition to Pop II star
formation, and here we give a brief overview of its development.
Considering only Pop III star formation, \citeauthor{Yoshida04}
calculated that Pop III stars can raise the mean metallicity to $-4.5
\lsim$ [Z/H]\footnote{We use the conventional notation, [Z/H] $\equiv
  \log(Z/H) - \log(Z_\odot/H_\odot)$.} $\lsim -3.5$ by redshift 15 in
the upper limit where all metal-free stars have $M = 200 \Ms$ and
produce PISNe.  Considering both Pop III and II star formation,
\citeauthor{Scannapieco03} found that the transition between the two
modes is a gradual process where both modes are coeval, confirmed by
most of the later works.  Because the host galaxies have small masses
and are subject to negative radiative feedback through photo-heating
and photo-dissociation, the minimum halo mass gradually increases with
the radiation background intensity, which
\citeauthor{Furlanetto05_Reion} found to delay metal enrichment and
place it closer to the epoch of reionization.  \citeauthor{Trenti09}
noted that underdense regions of the universe are still pristine at
$z=6$ with Pop III stars still forming at these late epochs.  This
group later expanded on these results to find that Pop II SFR becomes
dominant at $z>25$ in the buildup of a MW-type halo.  Furthermore they
stress the importance of a photo-dissociating radiation background
that reduces enrichment by PISNe and increases the importance of
metal-rich galactic outflows, where only $10^{-4} - 10^{-2}$ of PISN
ejecta ends up in EMP stars with $M > 0.9 \Ms$.

The Pop III IMF should play an important role in abundance patterns in
extremely metal poor (EMP) stars in the Milky Way (MW) halo and the
physics of reionization.  Using these data as constraints,
\citeauthor{Tumlinson06} found that Pop III IMFs with log-normal
distributions with mean masses between 8 and 42 \Ms~best fit the data.
Furthermore he concludes that EMP stars with [Z/H] $<$ --3 have
between 1 and 10 metal-free SN progenitors and the Pop III SFR is less
than 1\% of the total SFR at $z=6$.  \citeauthor{Karlsson08} use
observational data of EMP stars with their model to constrain the mass
fraction of Pop III stars that die with a PISN is less than 40 per
cent.  They also conclude that stars enriched primarily by PISNe have
[Ca/H] $\gsim$ --2.6, which could have been missed by some EMP
surveys.

@@

Cosmological simulations are useful but computationally expensive to
detangle and study the transition from Pop III to II star formation,
especially the complexities ensuing from hierarchical structure
formation, stellar feedback, and inhomogeneous chemical enrichment.
\citeauthor{Tornatore07} employed smoothed particle hydrodynamics
(SPH) simulations to study this problem and found that Pop III star
formation occurs down to $z=2.5$ in underdense regions.  Their mass
resolution did not allow them to capture star formation in minihalos
($M_{\rm halo} \sim 10^{5-7} \Ms$) that host Pop III stars, which may
have resulted in an underestimate of metal enrichment.  They found a
`Pop III wave' of star formation, where Pop III star formation is
quenched in an inside-out fashion from biased regions to voids, is
however still valid.  \citeauthor{Ricotti08} used a Lagrangian
moving-mesh simulation with radiative, mechanical, and chemical
feedback that resolved minihaloes to find that (i) 1--10\% of the
cosmic volume is chemically enriched, (ii) the gas reservoir in halos
with $M < 10^8 \Ms$ are depleted, (iii) there is a large scatter in
the mass-to-light (M/L) ratios of dwarf galaxies, and (iv) their
luminosity function is relatively flat.  Most recently
\citeauthor{Maio10_Pop32} found that the ratios of Pop III to Pop II
SFRs are insensitive to the exact value of $Z_{\rm cr}$.  In contrast
with \citeauthor{Ricotti08}, only $10^{-8}$ of the simulation volume
is chemically enriched by $z=11$.  This difference of several orders
of magnitude may result from the lack of radiative feedback, which can
expel most of the gas from the host halo before the SN explosion.
Supporting this idea, \citet{Whalen08_SN} found that PISN blastwaves
in neutral halos radiate all of their energy, stalling the shock
within the halo and suppressing outflows.

Here, we present simulations that include both modes of star formation
and their radiative, mechanical, and chemical feedback, extending the
aforementioned previous work on the transition from Pop III to II star
formation and the birth of the first galaxies.  The methods used here
incorporate and link together recent results from metal-enriched and
metal-free star formation, the critical metallicity, and
pair-instability supernovae.  There are some uncertainities in these
input parameters, and we thus run several simulations while varying
these parameters.  This is the first time it has been possible to link
the formation and feedback of the first stars to protogalaxies in a
numerical simulation, resolving the important scales and including the
most important physical effects.

This paper is structured as follows.  Section \ref{sec:setup}
describes our simulation setup, numerical methods for star formation
and feedback, and different physical models considered in each
simulation.  Next we present our results on the stellar populations
and metal enrichement of the high-redshift dwarf galaxies in Section
\ref{sec:results}.  Then in Section \ref{sec:models}, we show the
changes in galactic properties and metal enrichments in each physical
model.  In Section \ref{sec:compare}, we compare our results with
previous work on the transition to Pop II.  We then discuss the
importance of Pop III star formation on high-redshift galaxy
formation and the implications on the interpretation of damped
Lyman-$\alpha$ absorbers (DLAs) and galactic archaeology in Section
\ref{sec:discuss}.  Finally in Section \ref{sec:summary}, we summarize
our results.

@@

\begin{figure*}
  \epsscale{2.0}
  \plotone{sfh1.eps}
  \epsscale{1.0}
  \caption{\label{fig:SFR} Star formation history.}
\end{figure*}
